Retrofitting Effect Handlers onto OCaml

### 2021-03-24 retro-concurrency.tex
Updates

The remainder of the paper continues with a description of the stock OCaml
program stack (\S\ref{sec:stack}). We describe effect handlers in Multicore
OCaml focussing on the challenges in retrofitting them into a mainstream
\emph{systems} language (\S\ref{sec:refine}), followed by the static and
dynamic semantics for Multicore OCaml effect handlers (\S\ref{sec:semantics}).
We then discuss the compiler and the runtime system support for implementing
effect handlers (\S\ref{sec:impl}), and present an extensive performance
evaluation of effect handlers (\S\ref{sec:eval}) against our design goals
(\S\ref{sec:req}). Finally, we discuss the related work (\S\ref{sec:related})
and conclude (\S\ref{sec:conc}).

@@

Logically, DWARF call-frame information maintains a large table which records
for every machine instruction where the return address and callee-saved
registers are stored. To avoid reifying this large table, DWARF directives
represent the table using a compact bytecode representation. Each block of such
instructions usually describe the call-frame information for a single function,
where each instruction describes the table as a delta of the state at the
previous instruction. While this leads to a compact representation, computing
the return address at a given instruction requires interpreting the bytecode
from the start of the function, which may be potentially very slow. Within each
function, DWARF maintains a \emph{canonical frame address} (CFA) and is
traditionally the stack pointer before entering this function. Hence, on
x86-64, where the return address is pushed on the stack on call, the return
address is at |CFA - 8|.

Our goal is to encode the CFA using DWARF directives when stacks are switched.
Recall that stack switching occurs in effect handlers, external calls and
callbacks. When starting to run on a new fiber, we insert DWARF bytecode to
follow the |parent_fiber| pointer and dereference the |saved_sp| to get the CFA
(|saved_sp + 8|). During callbacks into OCaml, we save the current system stack
pointer in the |context block| in Figure~\ref{fig:fiber} to identify the CFA in
the C stack. DWARF unwinding for external calls is implemented by following a
link to the current OCaml stack pointer. With these changes, we get the same
backtrace for the \texttt{meander} program from \S\ref{sec:unwind}, modulo
runtime system functions due to effect handlers.

@@

Despite the correct DWARF unwind information, using DWARF to record call stack
information in |perf| only captures the call stack of the current fiber in
Multicore OCaml. Since stack unwinding using DWARF is slow due to bytecode
interpretation overhead, |perf| dumps the (user) call stack when
sampled~\cite{Bastian19}. This only includes the frames from the current fiber.
This is a limitation of |perf| and not of our stack layout. Bastian et
al.~\cite{Bastian19} report on a technique to pre-compile the unwind table to
assembly which speeds up DWARF-based unwinding by upto 25$\times$. With this
fast unwinding, |perf| would no longer needs to dump the call stack, and hence,
we can get the backtrace of the entire call stack and not just the current
fiber.

@@

This research was funded via a Royal Commission for the Exhibition
of 1851 and Darwin College Research Fellowships, and by grants from Jane Street
and Tezos Foundation.

>>>>

The remainder of the paper continues with a description of the stock OCaml
program stack (\S\ref{sec:stack}). We then describe effect handlers in
Multicore OCaml focussing on the challenges in retrofitting them into a
mainstream \emph{systems} language (\S\ref{sec:refine}), followed by the static
and dynamic semantics for Multicore OCaml effect handlers
(\S\ref{sec:semantics}). We then discuss the compiler and the runtime system
support for implementing effect handlers (\S\ref{sec:impl}), and present an
extensive performance evaluation of effect handlers (\S\ref{sec:eval}) against
our design goals (\S\ref{sec:req}). Finally, we discuss the related work
(\S\ref{sec:related}) and conclude (\S\ref{sec:conc}).

@@

Logically, DWARF call-frame information maintains a large table which records
for every machine instruction where the return address and callee-saved
registers are stored. To avoid reifying this large table, DWARF directives
represent the table using a compact bytecode representation. Each block of such
instructions usually describe the call-frame information for a single function,
where each instruction describes the table as a delta of the state at the
previous instruction. While this leads to a compact representation, computing
the return address at a given instruction requires interpreting the bytecode
from the start of the function, which may potentially be very slow.

Within each function, DWARF maintains a \emph{canonical frame address} (CFA)
and is traditionally the stack pointer before entering this function. Hence, on
x86-64, where the return address is pushed on the stack on call, the return
address is at |CFA - 8|. Our goal is to encode the CFA using DWARF directives
when stacks are switched. Recall that stack switching occurs in effect
handlers, external calls and callbacks. When starting to run on a new fiber, we
insert DWARF bytecode to follow the |parent_fiber| pointer and dereference the
|saved_sp| to get the CFA (|saved_sp + 8|). During callbacks into OCaml, we
save the current system stack pointer in the |context block| in
Figure~\ref{fig:fiber} to identify the CFA in the C stack. DWARF unwinding for
external calls is implemented by following a link to the current OCaml stack
pointer. With these changes, we get the same backtrace for the \texttt{meander}
program from \S\ref{sec:unwind}, modulo runtime system functions due to effect
handlers.

@@

Despite the correct DWARF unwind information, using DWARF to record call stack
information in |perf| only captures the call stack of the current fiber in
Multicore OCaml. Since stack unwinding using DWARF is slow due to bytecode
interpretation overhead, |perf| dumps the (user) call stack when
sampled~\cite{Bastian19}. This only includes the frames from the current fiber.
This is a limitation of |perf| and not of our stack layout. Bastian et
al.~\cite{Bastian19} report on a technique to pre-compile the unwind table to
assembly which speeds up DWARF-based unwinding by upto 25$\times$. With this
fast unwinding, |perf| would no longer needs to dump the call stack, and hence,
we could get the backtrace of the entire call stack and not just the current
fiber.

@@

We thank our shepherd, Matthew Fluet, whose comments substantially improved the
presentation. This research was funded via a Royal Commission for the
Exhibition of 1851 and Darwin College Research Fellowships, and by grants from
Jane Street and Tezos Foundation.
